---
title: "Shiny apps with Databricks data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny apps with Databricks data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

brickster can query Databricks SQL warehouses, connect through DBI and dbplyr,
browse Unity Catalog, and read files from volumes. You can use these functions
in a Shiny app with a regular connection, such as the app's service principal.
In that case, requests use the permissions of that connection.

The Shiny helpers described here instead use each signed-in user's Databricks
authorization. Warehouse permissions, Unity Catalog grants, row filters, and
column masks determine what the user can access. Your app can use the permissions
you already manage in Databricks without maintaining another set of data-access
rules in Shiny.

shinyOAuth handles sign-in and token refresh. brickster supplies the Databricks
provider and connects those credentials to its SQL, DBI, and REST functions.
The example below runs data operations in background R processes with Shiny's
`ExtendedTask` and mirai, so the app stays responsive while a query runs.

## Setup

### Packages

Use a version of brickster containing the Shiny helpers. The helpers currently
need development shinyOAuth, because CRAN 0.6.1 does not yet provide its
`connection()` and `access_token()` APIs. Install shinyOAuth from GitHub:

```{r}
install.packages(c("pak", "shiny", "bslib", "promises", "mirai"))
pak::pak("lukakoning/shinyOAuth")
```

When using a brickster source checkout, install it before launching the app so
both the Shiny process and its background workers can load the package:

```{r}
pak::local_install(".")
```

Use Shiny 1.8.1 or later. brickster's SQL backend uses the Databricks Statement
Execution API; this example does not require an ODBC driver.

### Register the app

Ask a Databricks account administrator to register a custom OAuth application
in the account console under Settings, App connections. Register the exact
callback URL, for example `http://localhost:8080/callback` for local development
or `https://your-app.example/callback` when deployed. Copy its client ID and,
for a confidential application, its client secret.

Enable the SQL scope for an app that only queries a warehouse. The example also
browses catalogs and reads a volume file, so enable All APIs for those options.
Identity scopes and `offline_access` allow sign-in and token refresh. See
[Databricks OAuth application setup](https://learn.microsoft.com/en-us/azure/databricks/integrations/enable-disable-oauth).

Each user needs access to the workspace and the resources used by the app.
For SQL, that includes `CAN USE` on the warehouse, `USE CATALOG`, `USE SCHEMA`,
and `SELECT` on the relevant tables or views. Databricks checks these permissions
when the user's query runs.

### App settings

Add the following settings to `.Renviron`, then restart R. Use the workspace
hostname and a SQL warehouse ID from its connection details:

```text
DATABRICKS_HOST=your-workspace.cloud.databricks.com
DATABRICKS_CLIENT_ID=your-registered-client-id
DATABRICKS_CLIENT_SECRET=your-registered-client-secret
DATABRICKS_WAREHOUSE_ID=your-warehouse-id
SHINY_OAUTH_REDIRECT_URI=http://localhost:8080/callback
```

For an OAuth application registered without a secret, omit
`DATABRICKS_CLIENT_SECRET`. The helpers use PKCE in both cases.

The app initially runs `SELECT current_user() AS user`, which lets you confirm
whose authorization is being used. To load your data, add an SQL query such as:

```text
APP_SQL=SELECT * FROM main.analytics.sales LIMIT 100
```

For the optional volume example, add the path to a CSV file the user can read:

```text
APP_VOLUME_FILE=/Volumes/main/analytics/files/sales.csv
```

## A complete app

Save this code as `app.R`. SQL is the first option; the others demonstrate DBI,
Unity Catalog, and volume access with the same user's authorization.
The Load data button shows when an operation is running and prevents repeated
submissions. Results and errors from an earlier login are hidden.

```{r full-app}
library(shiny)
library(brickster)

mirai::daemons(2)
onStop(function() mirai::daemons(0))

secret <- Sys.getenv("DATABRICKS_CLIENT_SECRET")
config <- db_shiny_config(
  host = Sys.getenv("DATABRICKS_HOST"),
  client_id = Sys.getenv("DATABRICKS_CLIENT_ID"),
  redirect_uri = Sys.getenv("SHINY_OAUTH_REDIRECT_URI"),
  client_secret = if (nzchar(secret)) secret else NULL,
  scopes = "all-apis"
)
warehouse <- Sys.getenv("DATABRICKS_WAREHOUSE_ID")
sql <- Sys.getenv("APP_SQL", "SELECT current_user() AS user")
volume_file <- Sys.getenv("APP_VOLUME_FILE")

read_data <- function(source, host, token, warehouse, sql, volume_file) {
  switch(
    source,
    sql = brickster::db_sql_query(
      warehouse_id = warehouse,
      statement = sql,
      host = host,
      token = token,
      disposition = "INLINE",
      row_limit = 100,
      show_progress = FALSE
    ),
    dbi = {
      con <- DBI::dbConnect(
        brickster::DatabricksSQL(),
        warehouse_id = warehouse,
        host = host,
        token = token,
        disposition = "INLINE",
        show_progress = FALSE
      )
      on.exit(DBI::dbDisconnect(con))
      DBI::dbGetQuery(con, sql, row_limit = 100)
    },
    catalogs = {
      catalogs <- brickster::db_list_all_pages(
        brickster::db_uc_catalogs_list,
        host = host,
        token = token
      )
      tibble::tibble(name = purrr::map_chr(catalogs, "name"))
    },
    volume = {
      if (!nzchar(volume_file)) {
        stop("Set APP_VOLUME_FILE to a CSV file in a volume.")
      }
      path <- fs::file_temp(ext = "csv")
      on.exit(fs::file_delete(path))
      brickster::db_volume_read(
        volume_file,
        path,
        host = host,
        token = token,
        progress = FALSE
      )
      utils::read.csv(path, nrows = 100)
    }
  )
}

ui <- db_shiny_ui(
  "auth",
  config,
  fluidPage(
    titlePanel("Databricks data"),
    actionButton("login", "Sign in"),
    actionButton("logout", "Sign out"),
    textOutput("user"),
    selectInput(
      "source",
      "Read data using",
      c(
        "SQL query" = "sql",
        "DBI query" = "dbi",
        "Unity Catalog" = "catalogs",
        "Volume CSV" = "volume"
      )
    ),
    bslib::input_task_button("load", "Load data"),
    tableOutput("data")
  )
)

server <- function(input, output, session) {
  auth <- db_shiny_server("auth", config, auto_redirect = FALSE)
  query <- db_shiny_task(auth, read_data, button = "load")
  loaded_source <- reactiveVal(NULL)

  observeEvent(input$login, auth$login(), ignoreInit = TRUE)
  observeEvent(input$logout, auth$logout(), ignoreInit = TRUE)
  observeEvent(
    input$load,
    {
      if (
        query$invoke(
          source = input$source,
          warehouse = warehouse,
          sql = sql,
          volume_file = volume_file
        )
      ) {
        loaded_source(input$source)
      }
    },
    ignoreInit = TRUE
  )

  output$user <- renderText({
    if (!auth$ready()) {
      return("Sign in to load your Databricks data.")
    }
    claims <- auth$identity()$id_token_claims
    paste("Signed in:", if (is.null(claims$email)) claims$sub else claims$email)
  })
  output$data <- renderTable({
    req(identical(loaded_source(), input$source))
    head(query$result(), 100)
  })
}

shinyApp(
  ui,
  server,
  uiPattern = ".*",
  options = list(host = "127.0.0.1", port = 8080)
)
```

Run the app with `shiny::runApp()` and open `http://localhost:8080/`. Sign in,
leave SQL query selected, and click Load data. Select another option to try its
data access method. The volume option needs the additional setting above.

`db_shiny_task()` uses Shiny's `ExtendedTask` and mirai to run `read_data()` in
a background R process. It supplies `host` and `token`; the other inputs are
passed through `query$invoke()`. `query$result()` returns the data for the
current login or reports its error. Custom worker functions use qualified
package calls and receive their inputs as arguments, as `read_data()` does.

The SQL and DBI examples use `INLINE` results for small tables. Use
`disposition = "EXTERNAL_LINKS"` for larger results. brickster fetches the
temporary result links without sending Databricks credentials to cloud storage.

## DBI and dbplyr

The worker above uses a fixed token for one operation. For synchronous code
running in the Shiny session, use the refreshable provider. A DBI connection
retains it, so later queries obtain current credentials:

```{r}
con <- DBI::dbConnect(
  DatabricksSQL(), warehouse_id = warehouse,
  host = auth$host, token = auth$token_provider(),
  disposition = "INLINE", show_progress = FALSE
)
on.exit(DBI::dbDisconnect(con))

sales <- dplyr::tbl(con, DBI::Id(catalog = "main", schema = "analytics", table = "sales"))
sales |>
  dplyr::group_by(region) |>
  dplyr::summarise(total = sum(amount, na.rm = TRUE)) |>
  dplyr::collect()
```

Place this code in a server-side helper function so `on.exit()` closes the
connection. Calls in the Shiny process are synchronous; use the `ExtendedTask`
pattern from the full app for queries that may take time. Create and close the
DBI connection inside the worker using the supplied fixed token.

For filters from Shiny inputs, pass query parameters rather than constructing
SQL text from the input:

```{r}
db_sql_query(
  warehouse_id = warehouse,
  statement = "SELECT * FROM main.analytics.sales WHERE region = :region LIMIT 100",
  parameters = list(list(name = "region", type = "STRING", value = region)),
  host = host, token = token, disposition = "INLINE", show_progress = FALSE
)
```

## Other Databricks APIs

The same `token` argument works with brickster's REST wrappers, for example
`db_uc_tables_list()` and `db_volume_read()`. Choose the API scopes
allowed by your OAuth registration and the permissions needed by your users.
An app that only uses SQL can leave `scopes = "sql"`, the configuration default,
and remove the catalog and volume options from the example.

Job submission is different from querying data directly: a Databricks job
executes as its configured Run as identity. A user's permission to trigger a
job does not make its tasks execute as that user. See
[Databricks job identities](https://docs.databricks.com/aws/en/jobs/privileges).

## Long-running queries

`db_sql_query()` submits a statement, waits for it, and returns a data frame.
For a query you need to resume after an interrupted status check or token
expiry, use the lower-level statement functions instead. Save the statement ID
as soon as submission returns; subsequent checks can obtain a fresh token and
continue the same statement.

For example, add these controls to the app's UI:

```{r long-query-ui}
bslib::input_task_button("submit", "Submit query")
bslib::input_task_button("check", "Check query")
verbatimTextOutput("submitted")
verbatimTextOutput("query_state")
```

Inside `server()`, after creating `auth`, use separate tasks for submission and
status checks:

```{r long-query-server}
submission <- db_shiny_task(auth, db_sql_exec_query, button = "submit")
poll <- db_shiny_task(auth, db_sql_exec_status, button = "check")
saved_statement <- reactiveVal(NULL)

observeEvent(auth$generation(), saved_statement(NULL), ignoreNULL = FALSE)
observeEvent(input$submit, {
  if (submission$invoke(
    statement = sql, warehouse_id = warehouse,
    wait_timeout = "0s", disposition = "INLINE", row_limit = 100
  )) {
    saved_statement(NULL)
  }
}, ignoreInit = TRUE)
observeEvent(submission$status(), {
  req(submission$status() == "success")
  saved_statement(list(
    id = submission$result()$statement_id,
    generation = auth$generation()
  ))
})
observeEvent(input$check, {
  saved <- saved_statement()
  req(saved, identical(saved$generation, auth$generation()))
  poll$invoke(statement_id = saved$id)
}, ignoreInit = TRUE)

output$submitted <- renderPrint(submission$result())
output$query_state <- renderPrint({
  saved <- saved_statement()
  req(saved, identical(saved$generation, auth$generation()))
  result <- poll$result()
  req(identical(result$statement_id, saved$id))
  result
})
```

Click Submit query once, then Check query to inspect its progress. Each check
acquires current credentials for the same login. A failed check leaves the
saved ID available, so you can check again without resubmitting. Signing out
clears the app's saved ID; a new login does not inherit the earlier query.

A successful status response includes the result manifest and first result
chunk. For subsequent chunks, create another task with
`db_shiny_task(auth, db_sql_exec_result)` and invoke it with
`statement_id = saved$id` and `chunk_index = result$result$next_chunk_index`,
using the same login checks. These functions return the API response lists;
`db_sql_query()` remains the simpler choice when you just need a data frame.

If submission itself fails before returning an ID, the app cannot tell whether
Databricks accepted the statement. Do not resubmit automatically.

## Refresh, background tasks, and deployment

`auth$token_provider()` belongs to the current login. Token refresh keeps that
login's `generation()` unchanged. Signing out or signing in again invalidates
previous providers; obtain a new provider after the next login.

For a background task, obtain the token asynchronously in the Shiny process,
then send the fixed token and ordinary query arguments to mirai. The example
checks `generation()` before showing results, so a result from an earlier login
does not appear after the user signs out or changes login.

`min_valid_for` is the minimum token lifetime needed to start an operation,
not a query timeout. Workers cannot refresh tokens. For queries that may outlast
a token, use the statement-ID workflow above. Signing out does not cancel SQL
already submitted to Databricks.

Set up mirai daemons once when the app starts. The example uses two workers;
adjust this to your host. `onStop()` closes them when the app stops. Database
connections should be created and closed in the process that uses them.

For deployment, register the app's exact HTTPS callback URL and use it in
`SHINY_OAUTH_REDIRECT_URI`. The `uiPattern = ".*"` setting lets Shiny route the
callback subpath. Refer to shinyOAuth's deployment guidance for your hosting
service's proxy and callback configuration. `reauth_after_seconds` defaults to
eight hours; adjust it through `db_shiny_server()` when needed.
